---
title: "Ace2 PCR results, Culex pipiens complex"
author: "Norah Saarman"
date: "2024-02-27"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## PCR results

```{r}
ace_results <- read.csv("VecTechAce2Results_27Feb2024.csv")
```

Allele frequencies for each locality

```{r}
table(ace_results$Species,ace_results$Locality)
```
Genotypes in p/q notation, pipiens = pp, quinquefasciatus = qq, 

```{r}
# genotype df
df_geno <- data.frame(sampleID = ace_results$SampleID, locality = paste0(substr(ace_results$SampleID,1,3),"-",substr(ace_results$SampleID,4,5)), genotype = ace_results$Species, latitude = ace_results$Latitude, longitude = ace_results$Longitude)

df_geno$genotype[df_geno$genotype == "Culex pipiens pipiens"] <- "pp"
df_geno$genotype[df_geno$genotype == "hybrid"] <- "pq"
df_geno$genotype[df_geno$genotype == "Culex quenquefasciatus"] <- "qq"

df_geno

table(df_geno[,2:3])
```

## Pie charts on a map

Install any packages you don't already have
install.packages("LEA")

Load all the packages you might need:

library(LEA)
library(fields)
library(RColorBrewer)
library(mapplots)
library(maptools)
library(raster)
library(rgdal)
library(maptools)
source("http://membres-timc.imag.fr/Olivier.Francois/Conversion.R")
source("http://membres-timc.imag.fr/Olivier.Francois/POPSutilities.R")


One pie chart at a time:
```{r}
tab <- table(df_geno[,2:3])

dimnames(tab)$locality

for (i in dimnames(tab)$locality){
  pie(tab[i,], main=i,col=c("black","grey","white"),radius=.8,clockwise=TRUE,labels = NA)
}
```
convert to frequencies
```{r}
table <- tab

#read in the sample proportion of each genotype into three columns
for (i in 1:length(tab[,1])){
  tab[i,] <- tab[i,]/sum(tab[i,])
}

tab
```


define the reaches of your map
```{r}
coord = unique(df_geno[,c(5,4)])
rownames(coord) <- unique(df_geno[,2])

plot(coord, xlab = "Longitude", ylab = "Latitude")

library(maps)
library(mapdata)
map("worldHires","USA")
map(add = T, col = "grey90", fill = TRUE)
points(coord,col="red",cex=0.6)

map('state',col = "grey90", fill = TRUE)
```


#add background map of the world.
map(add = T, col = "grey90", fill = TRUE)




#add in pie charts
for (i in 1:33){
  add.pie(z = qpop[i,],coord.pop[i,3], x = coord.pop[i,1], y = coord.pop[i,2], clockwise=TRUE, labels = "", col = c("red","yellow"))
}
```
