---
title: "Geographic overlap from SDM and Ace2 results"
author: "Norah Saarman"
date: "2025-04-18"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# Load libraries
library(raster)
library(terra)
library(sf)
library(dplyr)
library(maps)
library(ggplot2)
library(sp)
library(mapdata)
library(fields)
library(mapplots)
library(clue)
```

# 1. Load and Process Species Distribution Models (SDMs)

```{r load-sdms}
# Load MaxEnt SDMs
sdm_pipiens <- raster("../gis/culex_pipiens_meansuitability.nc")
sdm_quinque <- raster("../gis/culex_quinquefasciatus_meansuitability.nc")

# Threshold to binary
threshold <- 0.5
sdm_pipiens_bin <- sdm_pipiens >= threshold
sdm_quinque_bin <- sdm_quinque >= threshold
```

# 2. Convert SDMs to Polygons

```{r convert-polygons}
# Raster to terra
sdm_pipiens_v <- terra::rast(sdm_pipiens_bin)
sdm_quinque_v <- terra::rast(sdm_quinque_bin)

# Raster to polygons
poly_pipiens <- terra::as.polygons(sdm_pipiens_v, dissolve = TRUE)
poly_quinque <- terra::as.polygons(sdm_quinque_v, dissolve = TRUE)

# Terra to sf
poly_pipiens_sf <- st_as_sf(poly_pipiens)
poly_quinque_sf <- st_as_sf(poly_quinque)

# Filter to presence only
names(poly_pipiens_sf)[1] <- "presence"
names(poly_quinque_sf)[1] <- "presence"
poly_pipiens_sf <- poly_pipiens_sf %>% filter(presence == 1)
poly_quinque_sf <- poly_quinque_sf %>% filter(presence == 1)
```

# 3. Project to Albers Equal Area and Calculate Overlap

```{r project-and-overlap}
# Define CRS
aea_crs <- st_crs("+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-96")

# Project
poly_pipiens_sf <- st_transform(poly_pipiens_sf, aea_crs)
poly_quinque_sf <- st_transform(poly_quinque_sf, aea_crs)

# Overlap
overlap_sf <- st_intersection(poly_pipiens_sf, poly_quinque_sf)

# Calculate areas (km²)
area_pipiens_km2 <- sum(st_area(poly_pipiens_sf)) / 1e6
area_quinque_km2 <- sum(st_area(poly_quinque_sf)) / 1e6
area_overlap_km2 <- sum(st_area(overlap_sf)) / 1e6
```

# 4. Clip to North America and Reproject to WGS84

```{r clip-and-reproject}
# US base map
us_states <- st_as_sf(map("usa", plot = FALSE, fill = TRUE))
us_states <- st_transform(us_states, crs = aea_crs)

# Clip extent
bbox_na <- st_as_sfc(st_bbox(c(xmin = -170, xmax = -50, ymin = 5, ymax = 85), crs = st_crs(4326)))
bbox_na_sf <- st_transform(bbox_na, crs = aea_crs)

# Clip
poly_pipiens_sf <- st_intersection(st_make_valid(poly_pipiens_sf), bbox_na_sf)
poly_quinque_sf <- st_intersection(st_make_valid(poly_quinque_sf), bbox_na_sf)
overlap_sf <- st_intersection(st_make_valid(overlap_sf), bbox_na_sf)

# Reproject to WGS84
poly_pipiens_ll <- st_transform(poly_pipiens_sf, 4326)
poly_quinque_ll <- st_transform(poly_quinque_sf, 4326)
overlap_ll <- st_transform(overlap_sf, 4326)
```

# 5. Visualize SDMs and Overlap

```{r plot-initial-overlap}
par(xpd = NA)
map("usa")
map(add = TRUE, col = "grey90", fill = TRUE)

plot(st_geometry(poly_pipiens_ll), col = rgb(0, 0, 1, 0.3), border = NA, add = TRUE)
plot(st_geometry(poly_quinque_ll), col = rgb(0, 1, 0, 0.3), border = NA, add = TRUE)
plot(st_geometry(overlap_ll), col = adjustcolor("#fc8d59", alpha.f = 1), border = NA, add = TRUE)

map("state", add = TRUE, col = "black", lwd = 0.5)
```

# 6. Merge Nearby Overlapping Regions

```{r merge-nearby-overlap}
# Reproject overlap
overlap_ll_proj <- st_transform(overlap_ll, aea_crs)

# Make valid and extract polygons
overlap_valid <- st_make_valid(overlap_ll_proj)
overlap_polygons <- st_collection_extract(overlap_valid, "POLYGON")
overlap_parts <- st_cast(overlap_polygons, "POLYGON")

# Buffer outward
buffer_dist_meters <- 25000
overlap_buffered <- st_buffer(overlap_parts, dist = buffer_dist_meters)
overlap_buffered <- st_make_valid(overlap_buffered)

# Merge touching patches
overlap_combined <- st_union(overlap_buffered)
overlap_combined <- st_make_valid(overlap_combined)

# Buffer inward
overlap_combined <- st_buffer(overlap_combined, dist = -buffer_dist_meters)
overlap_combined <- st_make_valid(overlap_combined)

# Finalize
overlap_combined <- st_cast(overlap_combined, "MULTIPOLYGON")
overlap_combined <- st_transform(overlap_combined, 4326)
```

# 7. Visualize Combined Overlap Regions

```{r plot-combined-overlap}
#pdf("../figs/overlap_SDM.pdf", width = 8, height = 6)
par(xpd = NA)
map("usa")
map(add = TRUE, col = "grey90", fill = TRUE)

plot(st_geometry(poly_pipiens_ll), col = rgb(0, 0, 1, 0.3), border = NA, add = TRUE)
plot(st_geometry(poly_quinque_ll), col = rgb(0, 1, 0, 0.3), border = NA, add = TRUE)
plot(st_geometry(overlap_combined), col = adjustcolor("#fc8d59", alpha.f = 0.9), border = NA, add = TRUE)

map("state", add = TRUE, col = "black", lwd = 0.5)
#dev.off()
```

# 8. Add Sampling Points and Population Structure (Pie Charts)

```{r}
# Load counts
counts <- read.csv("../data/ThisStudy.txt", sep = "\t")
coord <- as.data.frame(counts[,c("long","lat")])

# Prepare
countsDf <- as.data.frame(counts[,c(2,1,4,5,6,9,10,7,8)])
names(countsDf) <- c("locality","site","pp","pq","qq","latitude","longitude","year","h_index")
rownames(countsDf) <- countsDf$site

# Frequencies
freqsDf <- t(apply(countsDf[,c("pp","pq","qq")], 1, function(row) row / sum(row)))
freqsDf <- as.matrix(freqsDf)

# Plot pies
par(xpd = NA)
plot_order <- rev(order(countsDf$h_index))

map("usa")
map(add = TRUE, col = "grey90", fill = TRUE)

for (i in plot_order) {
  add.pie(z = freqsDf[i,],
          x = coord[i,1],
          y = coord[i,2],
          clockwise = TRUE,
          labels = "",
          col = c("black", "grey", "white"),
          cex = 1.5, radius = 1.5)
}
```
# 8. Visualize Combined Overlap Regions and pie charts together

```{r}
par(xpd = NA)
map("usa")
map(add = TRUE, col = "grey90", fill = TRUE)

plot(st_geometry(poly_pipiens_ll), col = rgb(0, 0, 1, 0.3), border = NA, add = TRUE)
plot(st_geometry(poly_quinque_ll), col = rgb(0, 1, 0, 0.3), border = NA, add = TRUE)
plot(st_geometry(overlap_combined), col = adjustcolor("#fc8d59", alpha.f = 0.9), border = NA, add = TRUE)

map("state", add = TRUE, col = "black", lwd = 0.5)

for (i in plot_order) {
  add.pie(z = freqsDf[i,],
          x = coord[i,1],
          y = coord[i,2],
          clockwise = TRUE,
          labels = "",
          col = c("black", "grey", "white"),
          cex = 1.5, radius = 1.5)
}
```

# 8. Visualize Combined Overlap Regions and Frame Pie Charts

```{r plot-combined-overlap-and-pies}
#pdf("../figs/ace2_pies_SDM.pdf", width = 8, height = 6)

# Expand x/y limits by a small margin (e.g., 2 degrees)
x_range <- range(coord[, 1] + jitterDf[, 1], na.rm = TRUE)
y_range <- range(coord[, 2] + jitterDf[, 2], na.rm = TRUE)

x_margin <- 8
y_margin <- 8

x_lim <- c(x_range[1] - x_margin, x_range[2] + x_margin)
y_lim <- c(y_range[1] - y_margin, y_range[2] + y_margin)

# Set xpd to NA to allow for plotting in the margins
par(xpd = NA)

# Plot map with expanded margins
map("usa", xlim = x_lim, ylim = y_lim)
map(add = TRUE, col = "grey90", fill = TRUE)

plot(st_geometry(poly_pipiens_ll), col = rgb(0, 0, 1, 0.3), border = NA, add = TRUE)
plot(st_geometry(poly_quinque_ll), col = rgb(0, 1, 0, 0.3), border = NA, add = TRUE)
plot(st_geometry(overlap_combined), col = adjustcolor("#fc8d59", alpha.f = 0.9), border = NA, add = TRUE)

map("state", add = TRUE, col = "black", lwd = 0.5)
# Allow plotting outside margins
par(xpd = NA)

# Extract USA map limits
usa_map <- map("usa", plot = FALSE)

# Calculate USA bounding box
x_min_map <- min(usa_map$x, na.rm = TRUE)
x_max_map <- max(usa_map$x, na.rm = TRUE)
y_min_map <- min(usa_map$y, na.rm = TRUE)
y_max_map <- max(usa_map$y, na.rm = TRUE)

# Set margins to expand outside the map
frame_margin_x <- 5  # longitude margin
frame_margin_y <- 5  # latitude margin

# Define expanded rectangle
x_min <- x_min_map - frame_margin_x
x_max <- x_max_map + frame_margin_x
y_min <- y_min_map - frame_margin_y
y_max <- y_max_map + frame_margin_y

# Number of pies you want manually:
n_top <- 8      # include corners
n_bottom <- 8   # include corners
n_left <- 4     # exclude corners
n_right <- 5    # exclude corners

# Create side pies:

# Top: 8 pies including corners
top_pies <- data.frame(
  longitude = seq(x_min, x_max, length.out = n_top),
  latitude = rep(y_max, n_top)
)

# Right: 5 pies excluding corners
right_pies <- data.frame(
  longitude = rep(x_max, n_right),
  latitude = seq(y_max, y_min, length.out = n_right + 2)[-c(1, n_right + 2)]  # drop corners
)

# Bottom: 8 pies including corners
bottom_pies <- data.frame(
  longitude = seq(x_max, x_min, length.out = n_bottom),
  latitude = rep(y_min, n_bottom)
)

# Left: 4 pies excluding corners
left_pies <- data.frame(
  longitude = rep(x_min, n_left),
  latitude = seq(y_min, y_max, length.out = n_left + 2)[-c(1, n_left + 2)]  # drop corners
)

# Stack pie coordinates clockwise
pie_coords <- rbind(top_pies, right_pies, bottom_pies, left_pies)

# Check how many pies total
n_pies_frame <- nrow(pie_coords)

# Number of pies needed = number of sites
n_sites <- length(plot_order)

# Check if number matches
if (n_sites > n_pies_frame) {
  stop("More sites than available pie positions! Increase pie counts.")
} else if (n_sites < n_pies_frame) {
  pie_coords <- pie_coords[1:n_sites, ]  # only use as many as needed
}

# Helper: globally optimal assignment
assign_sites_optimally <- function(pie_coords, site_coords, site_ids) {
  colnames(pie_coords) <- c("longitude", "latitude")
  colnames(site_coords) <- c("longitude", "latitude")
  
  dist_matrix <- as.matrix(dist(rbind(pie_coords, site_coords)))[
    1:nrow(pie_coords), (nrow(pie_coords)+1):(nrow(pie_coords)+nrow(site_coords))
  ]
  
  assignment <- solve_LSAP(dist_matrix)
  assigned_sites <- site_ids[assignment]
  return(assigned_sites)
}

# Assign pies optimally
site_coords_all <- coord[plot_order, ]
assignments <- assign_sites_optimally(pie_coords, site_coords_all, plot_order)

# Draw connecting lines and pies
for (i in seq_along(assignments)) {
  site_i <- assignments[i]
  
  x0 <- coord[site_i, "long"]
  y0 <- coord[site_i, "lat"]
  
  x1 <- pie_coords[i, "longitude"]
  y1 <- pie_coords[i, "latitude"]
  
  segments(x0 = x0, y0 = y0, x1 = x1, y1 = y1, col = "black", lty = 1, lwd = 1)
  
  add.pie(z = freqsDf[site_i, ],
          x = x1,
          y = y1,
          clockwise = TRUE,
          labels = "",
          col = c("black", "grey", "white"),
          cex = 1.5, radius = 1.5)
}
#dev.off
```