---
title: "Historical and Ace2 PCR results, Culex pipiens complex"
author: "Norah Saarman"
date: "2024-02-27"
output: pdf_document
---

R build: Geospatial 4.4.0 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(maps)
library(sp)
library(mapdata)
source("http://membres-timc.imag.fr/Olivier.Francois/POPSutilities.R")
library(fields)
library(mapplots)
```

## Import species/hybrid counts per site

```{r}
counts <- read.csv("../data/Barr1957_plusLit_plusThisStudy.txt", sep = "\t")
countsSp <- SpatialPoints(coords = cbind(counts$long,counts$lat))
```
Genotypes in p/q notation:
Cx. pipiens = pp
Cx. quinquefasciatus = qq


## Create a dataframe of counts
```{r}
# start data frame and name fields
countsDf <- as.data.frame(counts[,c(2,1,4,5,6,9,10,7,8)])
names(countsDf) <- c("locality","site","pp","pq","qq","latitude","longitude","year","h_index")

# name rows
rownames(countsDf)<- countsDf$site
```

## Pie charts on a map
Convert counts to proportions (frequency):
```{r}
freqsDf <- as.data.frame(countsDf[,c("pp","pq","qq")])
freqsDf <- as.matrix.data.frame(t(apply(freqsDf, 1, function(row) row / sum(row))))
```

One pie chart at a time, to check code
```{r, include=F }
for (i in 1:length(freqsDf[,1])){
  pie(freqsDf[i,],main=i,col=c("black","grey","white"),radius=.2,clockwise=TRUE,labels = NA)
}
```

Plot points on map to check data:
```{r}
# Set xpd to NA to allow for plotting in the margins
par(xpd = NA)

#create and plot coord = long, lat
coord <- as.data.frame(countsDf[,c("longitude","latitude")])

#plot coordinates onto map
map("usa")
map(add = T, col = "grey90", fill = TRUE)
points(coord,col="black",cex=1)
```

Add pies to map following: "http://membres-timc.imag.fr/Olivier.Francois/Conversion.R"

```{r}
# Open PDF device
#pdf("../figs/ace2_pies_BarrLitThis.pdf", width = 8, height = 6)

# Set xpd to NA to allow for plotting in the margins
par(xpd = NA)

# Determine plot order by descending h_index
plot_order <- rev(order(countsDf$h_index))

# plot pies onto map
map("usa")
map(add = T, col = "grey90", fill = TRUE)
for (i in plot_order){
  add.pie(z = freqsDf[i,], 
          x = coord[i,1], 
          y = coord[i,2], 
          clockwise=TRUE, 
          labels = "", 
          col = c("black","grey","white"), 
          cex = 1, radius = 1 )
}
#dev.off()
```
```{r, eval=F, include = FALSE}
# Add random jitter, but i don't really like this.

# Set xpd to NA to allow for plotting in the margins
par(xpd = NA)

# Plot base map
map("usa")
map(add = TRUE, col = "grey90", fill = TRUE)

# Jitter settings
x_jitter <- 2  # degrees longitude
y_jitter <- 2  # degrees latitude

set.seed(42)  # for reproducibility

for (i in plot_order) {
  # Jittered coordinates
  jittered_x <- coord[i, 1] + runif(1, -x_jitter, x_jitter)
  jittered_y <- coord[i, 2] + runif(1, -y_jitter, y_jitter)
  
  # Plot pie chart at jittered location
  add.pie(z = freqsDf[i,],
          x = jittered_x,
          y = jittered_y,
          clockwise = TRUE,
          labels = "",
          col = c("black", "grey", "white"),
          cex = 1.2, radius = 1.2)
}
```


Plot latitude vs h-index excluding 0s (pure quinq) and 1s (pure pip):
```{r}
library(ggplot2)

# Choose to include pure or not
df <- countsDf[countsDf$h_index>0 & countsDf$h_index<1,]
#df <- countsDf

library(dplyr)

df <- df %>%
  mutate(
    period = case_when(
      year >= 1940 & year <= 1960 ~ "1940–1960",
      year >= 1990 & year <= 1999 ~ "1990–1999",
      year >= 2000 & year <= 2009 ~ "2000–2009",
      year >= 2010 & year <= 2019 ~ "2010–2019",
      year >= 2020 & year <= 2024 ~ "2020–2024",
      TRUE ~ "Other"  # fallback for unexpected years
    )
  )

# Optional: custom colors to match the legend in your image
period_colors <- c(
  "1940–1960" = "#d73027",  # red
  "1990–1999" = "#fc8d59",  # orange
  "2000–2009" = "#e6b800",  # yellowish
  "2010–2019" = "#66c2a5",  # greenish teal
  "2020–2024" = "#1f78b4"   # blue
)

p <- ggplot(df, aes(x = h_index, y = latitude, color = period)) +
  geom_point(size = 3) +
  scale_color_manual(values = period_colors) +
  theme_minimal() +
  labs(x = "H-index", y = "Latitude", color = "Period")

p
# Save to PDF
#ggsave("../figs/hindex_latitude_by_period.pdf", plot = p, width = 8, height = 6)
```
Add a line of best fit per period
```{r}
library(ggplot2)
library(dplyr)

# Filter groups with more than 4 data points
df_filtered <- df %>%
  group_by(period) %>%
  filter(n() > 4) %>%
  ungroup()

p2 <- ggplot(df, aes(x = h_index, y = latitude, color = period)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(data = df_filtered, method = "lm", formula = y ~ x, se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = c(
    "1940–1960" = "#d73027",
    "1990–1999" = "#fc8d59",
    "2000–2009" = "#e6b800",
    "2010–2019" = "#66c2a5",
    "2020–2024" = "#1f78b4"
  )) +
  theme_minimal() +
  labs(x = "H-index", y = "Latitude", color = "Period")

p2
```


```{r}
library(ggplot2)
library(dplyr)
library(patchwork)  # for layout

# Full dataset
df_all <- df

# West/East of Mississippi (arbitrary cutoff: longitude < -90)
#df_west <- df %>% filter(longitude < -90)
#df_east <- df %>% filter(longitude >= -90)

# West/East of the Rockies (longitude < -105)
df_west <- df %>% filter(longitude < -105)
df_east <- df %>% filter(longitude >= -105)

# Filtered versions for trend lines
df_all_fit <- df_all %>% group_by(period) %>% filter(n() > 4) %>% ungroup()
df_west_fit <- df_west %>% group_by(period) %>% filter(n() > 4) %>% ungroup()
df_east_fit <- df_east %>% group_by(period) %>% filter(n() > 4) %>% ungroup()

# Define color palette
my_colors <- c(
  "1940–1960" = "#d73027",
  "1990–1999" = "#fc8d59",
  "2000–2009" = "#e6b800",
  "2010–2019" = "#66c2a5",
  "2020–2024" = "#1f78b4"
)

# Top panel: all data
p_all <- ggplot(df_all, aes(x = h_index, y = latitude, color = period)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(data = df_all_fit, method = "lm", formula = y ~ x, se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = my_colors) +
  theme_minimal() +
  ylim(30, 47) +
  labs(title = "All Data", x = "H-index", y = "Latitude", color = "Period")

# Bottom left: west
p_west <- ggplot(df_west, aes(x = h_index, y = latitude, color = period)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(data = df_west_fit, method = "lm", formula = y ~ x, se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = my_colors) +
  theme_minimal() +
  ylim(30, 47) +
  labs(title = "West of the Rockies", x = "H-index", y = "Latitude", color = NULL)

# Bottom right: east
p_east <- ggplot(df_east, aes(x = h_index, y = latitude, color = period)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(data = df_east_fit, method = "lm", formula = y ~ x, se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = my_colors) +
  theme_minimal() +
  ylim(30, 47) +
  labs(title = "East of the Rockies", x = "H-index", y = "Latitude", color = NULL)

# Combine using patchwork
(p_all + p_west + p_east) +
  plot_layout(design = "
  AA
  AA
  AA
  BC
  ")


```

# NEON ecoclimate Domains

```{r}
# Load required packages
library(sf)
library(dplyr)
library(ggplot2)

# Read the NEON ecoclimate domain shapefile
neon_domains <- st_read("~/git/ace2/gis/NEON_Domains.shp")

names(neon_domains)

# define df and Create 'period' column from 'year'
#df <- countsDf[countsDf$h_index>0 & countsDf$h_index<1,]
df <- countsDf

# Create period column
df <- df %>%
  mutate(
    period = case_when(
      year >= 1940 & year <= 1960 ~ "1940–1960",
      year >= 1990 & year <= 2019 ~ "1990–2019",
      year >= 2020 & year <= 2024 ~ "2020–2024",
      TRUE ~ "Other"
    )
  )

# Convert to sf and join
df_sf <- st_as_sf(df, coords = c("longitude", "latitude"), crs = 4326)
df_sf <- st_transform(df_sf, st_crs(neon_domains))
df_joined <- st_join(df_sf, neon_domains)

# Extract lat/lon for plotting
df_joined <- df_joined %>%
  mutate(
    longitude = st_coordinates(.)[,1],
    latitude = st_coordinates(.)[,2]
  )

# Plot
ggplot(df_joined, aes(x = h_index, y = latitude, color = period)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(data = df_east_fit, method = "lm", formula = y ~ x, se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = c(
    "1940–1960" = "#d73027",
    "1990–2019" = "#e6b800",
    "2020–2024" = "#1f78b4"
  )) +
  facet_wrap(~ domainName) +
  theme_minimal() +
  ylim(30, 47) +
  labs(x = "H-index", y = "Latitude", color = "Period")
```