---
title: "Ace2 PCR results, Culex pipiens complex"
author: "Norah Saarman"
date: "2024-02-27"
output: pdf_document
---

R build: Geospatial 4.4.0 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(maps)
library(sp)
library(mapdata)
source("http://membres-timc.imag.fr/Olivier.Francois/POPSutilities.R")
library(fields)
library(mapplots)
```

## Import species/hybrid counts per site

```{r}
counts <- read.csv("../data/ThisStudy.txt", sep = "\t")
countsSp <- SpatialPoints(coords = cbind(counts$long,counts$lat))
```
Genotypes in p/q notation:
Cx. pipiens = pp
Cx. quinquefasciatus = qq


## Create a dataframe of counts
```{r}
# start data frame and name fields
countsDf <- as.data.frame(counts[,c(2,1,4,5,6,9,10,7,8)])
names(countsDf) <- c("locality","site","pp","pq","qq","latitude","longitude","year","h_index")

# name rows
rownames(countsDf)<- countsDf$site
```

## Pie charts on a map
Convert counts to proportions (frequency):
```{r}
freqsDf <- as.data.frame(countsDf[,c("pp","pq","qq")])
freqsDf <- as.matrix.data.frame(t(apply(freqsDf, 1, function(row) row / sum(row))))
```

One pie chart at a time, to check code
```{r, include=F }
for (i in 1:length(freqsDf[,1])){
  pie(freqsDf[i,],main=i,col=c("black","grey","white"),radius=.2,clockwise=TRUE,labels = NA)
}
```

Plot points on map to check data:
```{r}
# Set xpd to NA to allow for plotting in the margins
#par(xpd = NA)

#create and plot coord = long, lat
coord <- as.data.frame(countsDf[,c("longitude","latitude")])

#plot coordinates onto map
map("usa") # world, usa, state, county, etc.
map(add = T, col = "grey90", fill = TRUE)
points(coord,col="black",cex=1)
```

Add pies to map following: "http://membres-timc.imag.fr/Olivier.Francois/Conversion.R"
```{r}
# Set xpd to NA to allow for plotting in the margins
par(xpd = NA)

# Determine plot order by descending h_index
plot_order <- rev(order(countsDf$h_index))

# plot pies onto map
map("usa")
map(add = T, col = "grey90", fill = TRUE)
for (i in plot_order){
  add.pie(z = freqsDf[i,], 
          x = coord[i,1], 
          y = coord[i,2], 
          clockwise=TRUE, 
          labels = "", 
          col = c("black","grey","white"), 
          cex = 1.5, radius = 1.5 )
}
```

```{r, eval=F, include = FALSE}
# Add random jitter, but i don't really like this.

# Set xpd to NA to allow for plotting in the margins
par(xpd = NA)

# Plot base map
map("usa")
map(add = TRUE, col = "grey90", fill = TRUE)

# Jitter settings
x_jitter <- 2  # degrees longitude
y_jitter <- 2  # degrees latitude

set.seed(42)  # for reproducibility

for (i in plot_order) {
  # Jittered coordinates
  jittered_x <- coord[i, 1] + runif(1, -x_jitter, x_jitter)
  jittered_y <- coord[i, 2] + runif(1, -y_jitter, y_jitter)
  
  # Plot pie chart at jittered location
  add.pie(z = freqsDf[i,],
          x = jittered_x,
          y = jittered_y,
          clockwise = TRUE,
          labels = "",
          col = c("black", "grey", "white"),
          cex = 1.2, radius = 1.2)
}
```
Create a jitter data frame by hand:
```{r}
# Create a jitter df
jitterDf <- data.frame(x_jitter = rep(0,25), y_jitter = rep(0,25))
rownames(jitterDf) <- rownames(countsDf)
jitterDf[3,] <- c(0,1)
jitterDf[4,] <- c(5,-1)
jitterDf[5,] <- c(1,-1)
jitterDf[6,] <- c(-2,0)
jitterDf[8,] <- c(2,-3)
jitterDf[9,] <- c(5,-1)
jitterDf[14,] <- c(-7,3.5)
jitterDf[16,] <- c(-6,-1.5)
jitterDf[19,] <- c(1,1)
jitterDf[20,] <- c(1,-1)
```
Calculate bounds of map based on coordinates + jitter: 
```{r}
# Expand x/y limits by a small margin (e.g., 2 degrees)
x_range <- range(coord[, 1] + jitterDf[, 1], na.rm = TRUE)
y_range <- range(coord[, 2] + jitterDf[, 2], na.rm = TRUE)

x_margin <- 2
y_margin <- 2

x_lim <- c(x_range[1] - x_margin, x_range[2] + x_margin)
y_lim <- c(y_range[1] - y_margin, y_range[2] + y_margin)
```
Add jitter with a line from true origin:
```{r}
# Open PDF device
#pdf("../figs/ace2_pies_ThisStudy.pdf", width = 8, height = 6)

# Set xpd to NA to allow for plotting in the margins
par(xpd = NA)

# Plot map with expanded margins
map("usa", xlim = x_lim, ylim = y_lim)
map(add = TRUE, col = "grey90", fill = TRUE)

for (i in plot_order) {
  # Jittered coordinates
  jittered_x <- coord[i, 1] + (jitterDf[i, 1])
  jittered_y <- coord[i, 2] + (jitterDf[i, 2])
  
  # Draw line from true location to jittered pie
  segments(x0 = coord[i, 1],
           y0 = coord[i, 2],
           x1 = jittered_x,
           y1 = jittered_y,
           col = "black", lty = 1, lwd = 1)
  
  # Plot pie chart at jittered location
  add.pie(z = freqsDf[i,],
          x = jittered_x,
          y = jittered_y,
          clockwise = TRUE,
          labels = "",
          col = c("black", "grey", "white"),
          cex = 1.5, radius = 1.5)
}

#dev.off()
```
```{r}
# Load required library
library(clue)

# Allow plotting outside margins
par(xpd = NA)

# Extract USA map limits
usa_map <- map("usa", plot = FALSE)

# Calculate USA bounding box
x_min_map <- min(usa_map$x, na.rm = TRUE)
x_max_map <- max(usa_map$x, na.rm = TRUE)
y_min_map <- min(usa_map$y, na.rm = TRUE)
y_max_map <- max(usa_map$y, na.rm = TRUE)

# Set margins to expand outside the map
frame_margin_x <- 5  # longitude margin
frame_margin_y <- 5  # latitude margin

# Define expanded rectangle
x_min <- x_min_map - frame_margin_x
x_max <- x_max_map + frame_margin_x
y_min <- y_min_map - frame_margin_y
y_max <- y_max_map + frame_margin_y

# Plot map
map("usa", xlim = c(x_min, x_max), ylim = c(y_min, y_max))
map(add = TRUE, col = "grey90", fill = TRUE)

# Number of pies you want manually:
n_top <- 8      # include corners
n_bottom <- 8   # include corners
n_left <- 4     # exclude corners
n_right <- 5    # exclude corners

# Create side pies:

# Top: 8 pies including corners
top_pies <- data.frame(
  longitude = seq(x_min, x_max, length.out = n_top),
  latitude = rep(y_max, n_top)
)

# Right: 5 pies excluding corners
right_pies <- data.frame(
  longitude = rep(x_max, n_right),
  latitude = seq(y_max, y_min, length.out = n_right + 2)[-c(1, n_right + 2)]  # drop corners
)

# Bottom: 8 pies including corners
bottom_pies <- data.frame(
  longitude = seq(x_max, x_min, length.out = n_bottom),
  latitude = rep(y_min, n_bottom)
)

# Left: 4 pies excluding corners
left_pies <- data.frame(
  longitude = rep(x_min, n_left),
  latitude = seq(y_min, y_max, length.out = n_left + 2)[-c(1, n_left + 2)]  # drop corners
)

# Stack pie coordinates clockwise
pie_coords <- rbind(top_pies, right_pies, bottom_pies, left_pies)

# Check how many pies total
n_pies_frame <- nrow(pie_coords)

# Number of pies needed = number of sites
n_sites <- length(plot_order)

# Check if number matches
if (n_sites > n_pies_frame) {
  stop("More sites than available pie positions! Increase pie counts.")
} else if (n_sites < n_pies_frame) {
  pie_coords <- pie_coords[1:n_sites, ]  # only use as many as needed
}

# Helper: globally optimal assignment
assign_sites_optimally <- function(pie_coords, site_coords, site_ids) {
  colnames(pie_coords) <- c("longitude", "latitude")
  colnames(site_coords) <- c("longitude", "latitude")
  
  dist_matrix <- as.matrix(dist(rbind(pie_coords, site_coords)))[
    1:nrow(pie_coords), (nrow(pie_coords)+1):(nrow(pie_coords)+nrow(site_coords))
  ]
  
  assignment <- solve_LSAP(dist_matrix)
  assigned_sites <- site_ids[assignment]
  return(assigned_sites)
}

# Assign pies optimally
site_coords_all <- coord[plot_order, ]
assignments <- assign_sites_optimally(pie_coords, site_coords_all, plot_order)

# Draw connecting lines and pies
for (i in seq_along(assignments)) {
  site_i <- assignments[i]
  
  x0 <- coord[site_i, "longitude"]
  y0 <- coord[site_i, "latitude"]
  
  x1 <- pie_coords[i, "longitude"]
  y1 <- pie_coords[i, "latitude"]
  
  # Line
  segments(x0 = x0, y0 = y0, x1 = x1, y1 = y1, col = "black", lty = 1, lwd = 1)
  
  # Pie
  add.pie(z = freqsDf[site_i, ],
          x = x1,
          y = y1,
          clockwise = TRUE,
          labels = "",
          col = c("black", "grey", "white"),
          cex = 1.5, radius = 1.5)
}

```