---
title: "Geographic overlap from SDM and Ace2 results"
author: "Norah Saarman"
date: "2025-04-18"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# Load libraries
library(raster)
library(terra)
library(sf)
library(dplyr)
library(maps)
library(ggplot2)
library(sp)
library(mapdata)
library(fields)
library(mapplots)
library(clue)
```

# 1. Load and Process Species Distribution Models (SDMs)

```{r load-sdms}
# Load MaxEnt SDMs
sdm_pipiens <- raster("../gis/culex_pipiens_meansuitability.nc")
sdm_quinque <- raster("../gis/culex_quinquefasciatus_meansuitability.nc")

# Threshold to binary
threshold <- 0.5
sdm_pipiens_bin <- sdm_pipiens >= threshold
sdm_quinque_bin <- sdm_quinque >= threshold
```

# 2. Convert SDMs to Polygons

```{r convert-polygons}
# Raster to terra
sdm_pipiens_v <- terra::rast(sdm_pipiens_bin)
sdm_quinque_v <- terra::rast(sdm_quinque_bin)

# Raster to polygons
poly_pipiens <- terra::as.polygons(sdm_pipiens_v, dissolve = TRUE)
poly_quinque <- terra::as.polygons(sdm_quinque_v, dissolve = TRUE)

# Terra to sf
poly_pipiens_sf <- st_as_sf(poly_pipiens)
poly_quinque_sf <- st_as_sf(poly_quinque)

# Filter to presence only
names(poly_pipiens_sf)[1] <- "presence"
names(poly_quinque_sf)[1] <- "presence"
poly_pipiens_sf <- poly_pipiens_sf %>% filter(presence == 1)
poly_quinque_sf <- poly_quinque_sf %>% filter(presence == 1)
```

# 3. Project to Albers Equal Area and Calculate Overlap

```{r project-and-overlap}
# Define CRS
aea_crs <- st_crs("+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-96")

# Project
poly_pipiens_sf <- st_transform(poly_pipiens_sf, aea_crs)
poly_quinque_sf <- st_transform(poly_quinque_sf, aea_crs)

# Overlap
overlap_sf <- st_intersection(poly_pipiens_sf, poly_quinque_sf)

# Calculate areas (km²)
area_pipiens_km2 <- sum(st_area(poly_pipiens_sf)) / 1e6
area_quinque_km2 <- sum(st_area(poly_quinque_sf)) / 1e6
area_overlap_km2 <- sum(st_area(overlap_sf)) / 1e6
```

# 4. Clip to North America and Reproject to WGS84

```{r clip-and-reproject}
# US base map
us_states <- st_as_sf(map("usa", plot = FALSE, fill = TRUE))
us_states <- st_transform(us_states, crs = aea_crs)

# Clip extent
bbox_na <- st_as_sfc(st_bbox(c(xmin = -170, xmax = -50, ymin = 5, ymax = 85), crs = st_crs(4326)))
bbox_na_sf <- st_transform(bbox_na, crs = aea_crs)

# Clip
poly_pipiens_sf <- st_intersection(st_make_valid(poly_pipiens_sf), bbox_na_sf)
poly_quinque_sf <- st_intersection(st_make_valid(poly_quinque_sf), bbox_na_sf)
overlap_sf <- st_intersection(st_make_valid(overlap_sf), bbox_na_sf)

# Reproject to WGS84
poly_pipiens_ll <- st_transform(poly_pipiens_sf, 4326)
poly_quinque_ll <- st_transform(poly_quinque_sf, 4326)
overlap_ll <- st_transform(overlap_sf, 4326)
```

# 5. Visualize SDMs and Overlap

```{r plot-initial-overlap}
par(xpd = NA)
map("usa")
map(add = TRUE, col = "grey90", fill = TRUE)

plot(st_geometry(poly_pipiens_ll), col = rgb(0, 0, 1, 0.3), border = NA, add = TRUE)
plot(st_geometry(poly_quinque_ll), col = rgb(0, 1, 0, 0.3), border = NA, add = TRUE)
plot(st_geometry(overlap_ll), col = adjustcolor("#fc8d59", alpha.f = 1), border = NA, add = TRUE)

map("state", add = TRUE, col = "black", lwd = 0.5)
```

# 6. Merge Nearby Overlapping Regions

```{r merge-nearby-overlap}
# Reproject overlap
overlap_ll_proj <- st_transform(overlap_ll, aea_crs)

# Make valid and extract polygons
overlap_valid <- st_make_valid(overlap_ll_proj)
overlap_polygons <- st_collection_extract(overlap_valid, "POLYGON")
overlap_parts <- st_cast(overlap_polygons, "POLYGON")

# Buffer outward
buffer_dist_meters <- 25000
overlap_buffered <- st_buffer(overlap_parts, dist = buffer_dist_meters)
overlap_buffered <- st_make_valid(overlap_buffered)

# Merge touching patches
overlap_combined <- st_union(overlap_buffered)
overlap_combined <- st_make_valid(overlap_combined)

# Buffer inward
overlap_combined <- st_buffer(overlap_combined, dist = -buffer_dist_meters)
overlap_combined <- st_make_valid(overlap_combined)

# Finalize
overlap_combined <- st_cast(overlap_combined, "MULTIPOLYGON")
overlap_combined <- st_transform(overlap_combined, 4326)
```

# 7. Visualize Combined Overlap Regions

```{r plot-combined-overlap}
#pdf("../figs/overlap_SDM.pdf", width = 8, height = 6)
par(xpd = NA)
map("usa")
map(add = TRUE, col = "grey90", fill = TRUE)

plot(st_geometry(poly_pipiens_ll), col = rgb(0, 0, 1, 0.3), border = NA, add = TRUE)
plot(st_geometry(poly_quinque_ll), col = rgb(0, 1, 0, 0.3), border = NA, add = TRUE)
plot(st_geometry(overlap_combined), col = adjustcolor("#fc8d59", alpha.f = 0.9), border = NA, add = TRUE)

map("state", add = TRUE, col = "black", lwd = 0.5)
#dev.off()
```

# 8. Add Sampling Points and Population Structure (Pie Charts)

```{r}
# Load counts
counts <- read.csv("../data/ThisStudy_v9.txt", sep = "\t")
coord <- as.data.frame(counts[,c("long","lat")])

# Prepare
countsDf <- as.data.frame(counts[,c(2,1,4,5,6,9,10,7,8)])
names(countsDf) <- c("locality","site","pp","pq","qq","latitude","longitude","year","h_index")
rownames(countsDf) <- countsDf$site

# Frequencies
freqsDf <- t(apply(countsDf[,c("pp","pq","qq")], 1, function(row) row / sum(row)))
freqsDf <- as.matrix(freqsDf)

# Plot pies
par(xpd = NA)
plot_order <- rev(order(countsDf$h_index))

map("usa")
map(add = TRUE, col = "grey90", fill = TRUE)

for (i in plot_order) {
  add.pie(z = freqsDf[i,],
          x = coord[i,1],
          y = coord[i,2],
          clockwise = TRUE,
          labels = "",
          col = c("black", "grey", "white"),
          cex = 1.5, radius = 1.5)
}
```
Some jitter, added by hand:

Create a jitter data frame by hand:
```{r}
# Create a jitter df
jitterDf <- data.frame(x_jitter = rep(0,25), y_jitter = rep(0,25))
rownames(jitterDf) <- rownames(countsDf)
jitterDf[3,] <- c(0,1)
jitterDf[4,] <- c(5,-1)
jitterDf[5,] <- c(1,-1)
jitterDf[6,] <- c(-2,0)
jitterDf[8,] <- c(2,-3)
jitterDf[9,] <- c(5,-1)
jitterDf[14,] <- c(-7,3.5)
jitterDf[16,] <- c(-6,-1.5)
jitterDf[19,] <- c(1,1)
jitterDf[20,] <- c(1,-1)
```
Calculate bounds of map based on coordinates + jitter: 
```{r}
# Expand x/y limits by a small margin (e.g., 2 degrees)
x_range <- range(coord[, 1] + jitterDf[, 1], na.rm = TRUE)
y_range <- range(coord[, 2] + jitterDf[, 2], na.rm = TRUE)

x_margin <- 2
y_margin <- 2

x_lim <- c(x_range[1] - x_margin, x_range[2] + x_margin)
y_lim <- c(y_range[1] - y_margin, y_range[2] + y_margin)
```
Add jitter with a line from true origin:
```{r}
# Open PDF device
#pdf("../figs/ace2_pies_ThisStudy.pdf", width = 8, height = 6)

# Set xpd to NA to allow for plotting in the margins
par(xpd = NA)

# Plot map with expanded margins
map("usa", xlim = x_lim, ylim = y_lim)
map(add = TRUE, col = "grey90", fill = TRUE)

for (i in plot_order) {
  # Jittered coordinates
  jittered_x <- coord[i, 1] + (jitterDf[i, 1])
  jittered_y <- coord[i, 2] + (jitterDf[i, 2])
  
  # Draw line from true location to jittered pie
  segments(x0 = coord[i, 1],
           y0 = coord[i, 2],
           x1 = jittered_x,
           y1 = jittered_y,
           col = "black", lty = 1, lwd = 1)
  
  # Plot pie chart at jittered location
  add.pie(z = freqsDf[i,],
          x = jittered_x,
          y = jittered_y,
          clockwise = TRUE,
          labels = "",
          col = c("black", "grey", "white"),
          cex = 1.5, radius = 1.5)
}

#dev.off()
```

Predicted level of overlapping suitable habitat?
```{r}
# ---------------------------------------------
# Quantify predicted overlap around each sample point
# ---------------------------------------------

# Step 1: Make sample points an sf object
samples_sf <- st_as_sf(countsDf, coords = c("longitude", "latitude"), crs = 4326)

# Step 2: Reproject sample points to Albers Equal Area CRS
samples_sf <- st_transform(samples_sf, crs = aea_crs)

# Step 3: Buffer each point by 30 km (30,000 meters)
sample_buffers <- st_buffer(samples_sf, dist = 30000)

# Step 4: Prepare overlap_combined layer in same CRS
overlap_combined_proj <- st_transform(overlap_combined, crs = aea_crs)

# Step 5: Calculate proportion of buffer area that overlaps
predicted_overlap_prop <- sapply(1:nrow(sample_buffers), function(i) {
  buf <- sample_buffers[i,]
  intersection <- st_intersection(buf, overlap_combined_proj)
  
  if (nrow(intersection) == 0) {
    # No overlap
    overlap_area <- 0
  } else {
    # Sum all overlapping parts
    overlap_area <- sum(st_area(intersection))
  }
  
  buffer_area <- st_area(buf)
  
  # Return proportion
  as.numeric(overlap_area) / as.numeric(buffer_area)
})

# Step 6: Add predicted overlap to countsDf
countsDf$predicted_overlap <- predicted_overlap_prop

# Step 7: Quick check
head(countsDf[, c("site", "h_index", "predicted_overlap")])
```
Now, add directionality with 0 = quinq, 1 = pip, to estimate "predicted_h_index"

```{r}
# ---------------------------------------------
# New: Predict directional hybrid index from habitat around each sampling site
# ---------------------------------------------

# Step 1: Prepare the pipiens-only and quinque-only polygons
# (We already have poly_pipiens_ll and poly_quinque_ll, but need to transform)
poly_pipiens_proj <- st_transform(poly_pipiens_ll, crs = aea_crs)
poly_quinque_proj <- st_transform(poly_quinque_ll, crs = aea_crs)
overlap_combined_proj <- st_transform(overlap_combined, crs = aea_crs)  # already done above

# Step 2: Create pip-only and quinque-only polygons (remove overlap area)
pip_only <- st_difference(poly_pipiens_proj, overlap_combined_proj)
quinque_only <- st_difference(poly_quinque_proj, overlap_combined_proj)

# Step 3: Calculate areas for each sample buffer
library(units)  # make sure units package is loaded

predicted_h_index <- sapply(1:nrow(sample_buffers), function(i) {
  buf <- sample_buffers[i, ]
  
  pip_intersect <- st_intersection(buf, pip_only)
  quinque_intersect <- st_intersection(buf, quinque_only)
  overlap_intersect <- st_intersection(buf, overlap_combined_proj)
  
  pip_area <- if (nrow(pip_intersect) == 0) units::set_units(0, "m^2") else sum(st_area(pip_intersect))
  quinque_area <- if (nrow(quinque_intersect) == 0) units::set_units(0, "m^2") else sum(st_area(quinque_intersect))
  overlap_area <- if (nrow(overlap_intersect) == 0) units::set_units(0, "m^2") else sum(st_area(overlap_intersect))
  
  total_area <- pip_area + quinque_area + overlap_area
  
  if (as.numeric(total_area) == 0) {
    return(NA)  # no habitat found
  } else {
    pred_h <- (pip_area + 0.5 * overlap_area) / total_area
    return(as.numeric(pred_h))  # strip units at the end
  }
})
# Step 4: Add predicted_h_index to countsDf
countsDf$predicted_h_index <- predicted_h_index

# Step 5: Quick check and save
(summary_table <- countsDf[, c("site", "latitude", "longitude", "h_index", "predicted_overlap", "predicted_h_index")])

# Save as CSV
write.csv(summary_table, file = "../data/summary_hindex_prediction.csv", row.names = FALSE)

```
Plot h_index versus latitude, and h_index versus predicted h_index based on SDM habitat MaxEnt models:
```{r}
library(ggplot2)
library(patchwork)

# ---------------------------------
# First plot: H-index vs Latitude
# ---------------------------------

# Fit linear model for H-index vs Latitude
lm_lat <- lm(latitude ~ h_index, data = countsDf)
r_squared_lat <- summary(lm_lat)$r.squared
r_squared_lat_text <- paste0("R² = ", round(r_squared_lat, 2))

p1 <- ggplot(countsDf, aes(x = h_index, y = latitude)) +
  geom_point(size = 3, alpha = 0.8, color = "black") +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "#1f78b4", linewidth = 1.2) +
  annotate("text", x = 0.05, y = max(countsDf$latitude, na.rm = TRUE) - 1, label = r_squared_lat_text, hjust = 0, size = 4) +
  theme_minimal() +
  labs(x = "H-index (Ace2)", y = "Latitude", title = "H-index vs Latitude") +
  xlim(0, 1)

# ---------------------------------
# Second plot: H-index vs Predicted H-index
# ---------------------------------

# Fit linear model for H-index vs Predicted H-index
lm_pred <- lm(predicted_h_index ~ h_index, data = countsDf)
r_squared_pred <- summary(lm_pred)$r.squared
r_squared_pred_text <- paste0("R² = ", round(r_squared_pred, 2))

p2 <- ggplot(countsDf, aes(x = h_index, y = predicted_h_index)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
  geom_point(size = 3, alpha = 0.8, color = "black") +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "#d73027", linewidth = 1.2) +
  annotate("text", x = 0.05, y = 0.95, label = r_squared_pred_text, hjust = 0, size = 4) +
  theme_minimal() +
  labs(x = "H-index (Ace2)", y = "Predicted H-index (from SDM)", 
       title = "Observed vs Predicted H-index") +
  xlim(0, 1) +
  ylim(0, 1)

# ---------------------------------
# Combine side-by-side
# ---------------------------------

(p1 | p2) +
  plot_layout(guides = 'collect')

```
HWE Fisher's Exact Tests
```{r}
library(dplyr)

# Create an empty list to store results
hwe_results <- list()

# Loop through each site
for (i in 1:nrow(countsDf)) {
  # Extract observed counts
  obs_pp <- countsDf$pp[i]
  obs_pq <- countsDf$pq[i]
  obs_qq <- countsDf$qq[i]
  
  total_genotypes <- obs_pp + obs_pq + obs_qq
  
  # Skip if total is 0
  if (total_genotypes == 0) {
    hwe_results[[countsDf$site[i]]] <- NA
    next
  }
  
  # Check if there are at least two non-zero genotype classes
  non_zero_classes <- sum(c(obs_pp, obs_pq, obs_qq) > 0)
  
  if (non_zero_classes < 2) {
    hwe_results[[countsDf$site[i]]] <- NA
    next
  }
  
  # Calculate allele frequencies
  p <- (2 * obs_pp + obs_pq) / (2 * total_genotypes)
  q <- 1 - p
  
  # Expected counts under HWE
  exp_pp <- p^2 * total_genotypes
  exp_pq <- 2 * p * q * total_genotypes
  exp_qq <- q^2 * total_genotypes
  
  # Round expected counts
  exp_pp <- round(exp_pp)
  exp_pq <- round(exp_pq)
  exp_qq <- round(exp_qq)
  
  # Create contingency table
  table_hwe <- matrix(
    c(obs_pp, obs_pq, obs_qq,
      exp_pp, exp_pq, exp_qq),
    nrow = 2, byrow = TRUE
  )
  colnames(table_hwe) <- c("pp", "pq", "qq")
  rownames(table_hwe) <- c("Observed", "Expected")
  
  # Perform Fisher's exact test
  fisher_test <- fisher.test(table_hwe, simulate.p.value = TRUE, B = 1e5)
  
  # Save the result
  hwe_results[[countsDf$site[i]]] <- list(
    p_value = fisher_test$p.value,
    observed = c(pp = obs_pp, pq = obs_pq, qq = obs_qq),
    expected = c(pp = exp_pp, pq = exp_pq, qq = exp_qq)
  )
}

# Convert to a dataframe
hwe_summary <- do.call(rbind, lapply(names(hwe_results), function(site) {
  res <- hwe_results[[site]]
  if (is.null(res) || all(is.na(res))) {
    return(data.frame(site = site, p_value = NA))
  } else {
    return(data.frame(site = site, p_value = res$p_value))
  }
}))

# View the result
hwe_summary

```

Test for Utah:
```{r}
# Subset  sites
sub_sites <- countsDf %>%
  filter(grepl("UT", site))  # or however you tag Utah sites

# Sum genotype counts
total_pp <- sum(sub_sites$pp)
total_pq <- sum(sub_sites$pq)
total_qq <- sum(sub_sites$qq)

# Now perform HWE test on (total_pp, total_pq, total_qq) as a pooled sample
# Total individuals
total_genotypes <- total_pp + total_pq + total_qq

# Allele frequencies
p <- (2 * total_pp + total_pq) / (2 * total_genotypes)
q <- 1 - p

# Expected counts
exp_pp <- round(p^2 * total_genotypes)
exp_pq <- round(2 * p * q * total_genotypes)
exp_qq <- round(q^2 * total_genotypes)

# Contingency table
table_hwe <- matrix(c(total_pp, total_pq, total_qq,
                      exp_pp, exp_pq, exp_qq),
                    nrow = 2, byrow = TRUE)
colnames(table_hwe) <- c("pp", "pq", "qq")
rownames(table_hwe) <- c("Observed", "Expected")

# Fisher's Exact Test
fisher_test_pool <- fisher.test(table_hwe, simulate.p.value = TRUE, B = 1e5)

# Print p-value
fisher_test_pool$p.value

```

Test for CA, OR, WA:
```{r}
# Subset sites
sub_sites <- countsDf %>%
  filter(grepl("CA|OR|WA", site))  # or however you tag Utah sites

# Sum genotype counts
total_pp <- sum(sub_sites$pp)
total_pq <- sum(sub_sites$pq)
total_qq <- sum(sub_sites$qq)

# Now perform HWE test on (total_pp, total_pq, total_qq) as a pooled sample
# Total individuals
total_genotypes <- total_pp + total_pq + total_qq

# Allele frequencies
p <- (2 * total_pp + total_pq) / (2 * total_genotypes)
q <- 1 - p

# Expected counts
exp_pp <- round(p^2 * total_genotypes)
exp_pq <- round(2 * p * q * total_genotypes)
exp_qq <- round(q^2 * total_genotypes)

# Contingency table
table_hwe <- matrix(c(total_pp, total_pq, total_qq,
                      exp_pp, exp_pq, exp_qq),
                    nrow = 2, byrow = TRUE)
colnames(table_hwe) <- c("pp", "pq", "qq")
rownames(table_hwe) <- c("Observed", "Expected")

# Fisher's Exact Test
fisher_test_pool <- fisher.test(table_hwe, simulate.p.value = TRUE, B = 1e5)

# Print p-value
fisher_test_pool$p.value

```

Eastern sites:
```{r}
# Subset sites
sub_sites <- countsDf %>%
  filter(grepl("NJ|MD|MA|DE|CT|NY|VA|NC|SC|GA|FL", site))

# Sum genotype counts
total_pp <- sum(sub_sites$pp)
total_pq <- sum(sub_sites$pq)
total_qq <- sum(sub_sites$qq)

# Now perform HWE test on (total_pp, total_pq, total_qq) as a pooled sample
# Total individuals
total_genotypes <- total_pp + total_pq + total_qq

# Allele frequencies
p <- (2 * total_pp + total_pq) / (2 * total_genotypes)
q <- 1 - p

# Expected counts
exp_pp <- round(p^2 * total_genotypes)
exp_pq <- round(2 * p * q * total_genotypes)
exp_qq <- round(q^2 * total_genotypes)

# Contingency table
table_hwe <- matrix(c(total_pp, total_pq, total_qq,
                      exp_pp, exp_pq, exp_qq),
                    nrow = 2, byrow = TRUE)
colnames(table_hwe) <- c("pp", "pq", "qq")
rownames(table_hwe) <- c("Observed", "Expected")

# Fisher's Exact Test
fisher_test_pool <- fisher.test(table_hwe, simulate.p.value = TRUE, B = 1e5)

# Print p-value
fisher_test_pool$p.value

```

Eastern sites:
```{r}
# Subset sites
sub_sites <- countsDf %>%
  filter(grepl("TX|NM|AZ", site))

# Sum genotype counts
total_pp <- sum(sub_sites$pp)
total_pq <- sum(sub_sites$pq)
total_qq <- sum(sub_sites$qq)

# Now perform HWE test on (total_pp, total_pq, total_qq) as a pooled sample
# Total individuals
total_genotypes <- total_pp + total_pq + total_qq

# Allele frequencies
p <- (2 * total_pp + total_pq) / (2 * total_genotypes)
q <- 1 - p

# Expected counts
exp_pp <- round(p^2 * total_genotypes)
exp_pq <- round(2 * p * q * total_genotypes)
exp_qq <- round(q^2 * total_genotypes)

# Contingency table
table_hwe <- matrix(c(total_pp, total_pq, total_qq,
                      exp_pp, exp_pq, exp_qq),
                    nrow = 2, byrow = TRUE)
colnames(table_hwe) <- c("pp", "pq", "qq")
rownames(table_hwe) <- c("Observed", "Expected")

# Fisher's Exact Test
fisher_test_pool <- fisher.test(table_hwe, simulate.p.value = TRUE, B = 1e5)

# Print p-value
fisher_test_pool$p.value
```

# Create a summary table

```{r}
# Merge HWE p-values into countsDf
countsDf$HWE_p_value <- hwe_summary$p_value[match(countsDf$site, hwe_summary$site)]

# Add total N per site
countsDf$N <- countsDf$pp + countsDf$pq + countsDf$qq

# Select and reorder columns
summary_sites <- countsDf %>%
  select(site, pp, pq, qq, N, latitude, longitude, h_index, predicted_overlap, predicted_h_index, HWE_p_value)

# Save to CSV
write.csv(summary_sites, file = "../data/summary_sites_hindex_overlap_HWE.csv", row.names = FALSE)
 
# Select and reorder columns
summary_sites <- countsDf %>%
  select(site, pp, pq, qq, N, latitude, longitude, h_index, predicted_overlap, predicted_h_index, HWE_p_value)

# Save to CSV
write.csv(summary_sites, file = "../data/summary_sites_hindex_overlap_HWE.csv", row.names = FALSE)

# Function to pool and run HWE
pool_hwe_test <- function(df_region) {
  total_pp <- sum(df_region$pp, na.rm = TRUE)
  total_pq <- sum(df_region$pq, na.rm = TRUE)
  total_qq <- sum(df_region$qq, na.rm = TRUE)
  
  total_N <- total_pp + total_pq + total_qq
  
  if (total_N == 0) {
    return(data.frame(total_pp, total_pq, total_qq, total_N, pooled_HWE_p_value = NA))
  }
  
  p <- (2 * total_pp + total_pq) / (2 * total_N)
  q <- 1 - p
  
  exp_pp <- round(p^2 * total_N)
  exp_pq <- round(2 * p * q * total_N)
  exp_qq <- round(q^2 * total_N)
  
  table_hwe <- matrix(c(total_pp, total_pq, total_qq,
                        exp_pp, exp_pq, exp_qq),
                      nrow = 2, byrow = TRUE)
  colnames(table_hwe) <- c("pp", "pq", "qq")
  rownames(table_hwe) <- c("Observed", "Expected")
  
  fisher_test <- fisher.test(table_hwe, simulate.p.value = TRUE, B = 1e5)
  
  return(data.frame(total_pp, total_pq, total_qq, total_N, pooled_HWE_p_value = fisher_test$p.value))
}

countsDf$region <- case_when(
  grepl("CA|OR|WA", countsDf$site) ~ "West",
  grepl("NJ|MD|MA|DE|CT|NY|VA|NC|SC|GA|FL", countsDf$site) ~ "East",
  grepl("TX|NM|AZ", countsDf$site) ~ "Southwest",
  TRUE ~ "Other"
)

# Apply by region
summary_regions <- countsDf %>%
  group_by(region) %>%
  group_modify(~ pool_hwe_test(.x)) %>%
  ungroup()

write.csv(summary_regions, file = "../data/summary_regions_hindex_overlap_HWE.csv", row.names = FALSE)
```