---
title: "Geographic overlap of species ranges from SDM"
author: "Norah Saarman"
date: "2025-04-16"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Working from rasters to make polygons of geographic overlap:
```{r}
# Load libraries
library(raster)
library(terra)
library(sf)
library(dplyr)
library(maps)
library(ggplot2)

# ----------------------------------------
# STEP 1: Load and threshold SDMs to binary
# ----------------------------------------

# Load MaxEnt SDMs as raster layers
sdm_pipiens <- raster("../gis/culex_pipiens_meansuitability.nc")
sdm_quinque <- raster("../gis/culex_quinquefasciatus_meansuitability.nc")

# Threshold to binary presence/absence using a 0.5 cutoff
threshold <- 0.5
sdm_pipiens_bin <- sdm_pipiens >= threshold
sdm_quinque_bin <- sdm_quinque >= threshold

# ----------------------------------------
# STEP 2: Convert binary rasters to sf polygons
# ----------------------------------------

# Convert binary raster to terra SpatRaster
sdm_pipiens_v <- terra::rast(sdm_pipiens_bin)
sdm_quinque_v <- terra::rast(sdm_quinque_bin)

# Convert to polygons (dissolving contiguous presence regions)
poly_pipiens <- terra::as.polygons(sdm_pipiens_v, dissolve = TRUE)
poly_quinque <- terra::as.polygons(sdm_quinque_v, dissolve = TRUE)

# Convert terra polygons to sf objects
poly_pipiens_sf <- st_as_sf(poly_pipiens)
poly_quinque_sf <- st_as_sf(poly_quinque)

# Rename value column to 'presence' and filter to presence == 1
names(poly_pipiens_sf)[1] <- "presence"
names(poly_quinque_sf)[1] <- "presence"

poly_pipiens_sf <- poly_pipiens_sf %>% filter(presence == 1)
poly_quinque_sf <- poly_quinque_sf %>% filter(presence == 1)

# ----------------------------------------
# STEP 3: Reproject to Albers Equal Area for area analysis
# ----------------------------------------

aea_crs <- st_crs("+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-96")

poly_pipiens_sf <- st_transform(poly_pipiens_sf, aea_crs)
poly_quinque_sf <- st_transform(poly_quinque_sf, aea_crs)

# ----------------------------------------
# STEP 4: Calculate overlap and areas
# ----------------------------------------

# Compute geographic overlap
overlap_sf <- st_intersection(poly_pipiens_sf, poly_quinque_sf)

# Calculate areas in km²
area_pipiens_km2 <- sum(st_area(poly_pipiens_sf)) / 1e6
area_quinque_km2 <- sum(st_area(poly_quinque_sf)) / 1e6
area_overlap_km2 <- sum(st_area(overlap_sf)) / 1e6

# ----------------------------------------
# STEP 5: Prepare clipping extent and US base map
# ----------------------------------------

# Load US state outlines and reproject to match polygons
us_states <- st_as_sf(map("usa", plot = FALSE, fill = TRUE))
us_states <- st_transform(us_states, crs = aea_crs)

# Define North America bounding box in WGS84
bbox_na <- st_as_sfc(st_bbox(c(
  xmin = -170, xmax = -50,
  ymin = 5, ymax = 85
), crs = st_crs(4326)))

# Reproject bbox to Albers for clipping
bbox_na_sf <- st_transform(bbox_na, crs = aea_crs)

# ----------------------------------------
# STEP 6: Clean geometries and clip to bbox
# ----------------------------------------

# Ensure all geometries are valid
poly_pipiens_sf <- st_make_valid(poly_pipiens_sf)
poly_quinque_sf <- st_make_valid(poly_quinque_sf)
overlap_sf <- st_make_valid(overlap_sf)
us_states <- st_make_valid(us_states)
bbox_na_sf <- st_make_valid(bbox_na_sf)

# Clip all geometries to North America extent
poly_pipiens_sf <- st_intersection(poly_pipiens_sf, bbox_na_sf)
poly_quinque_sf <- st_intersection(poly_quinque_sf, bbox_na_sf)
overlap_sf <- st_intersection(overlap_sf, bbox_na_sf)
us_states <- st_intersection(us_states, bbox_na_sf)

# ----------------------------------------
# STEP 7: Reproject to WGS84 for base R map plotting
# ----------------------------------------

poly_pipiens_ll <- st_transform(poly_pipiens_sf, crs = 4326)
poly_quinque_ll <- st_transform(poly_quinque_sf, crs = 4326)
overlap_ll <- st_transform(overlap_sf, crs = 4326)

```

## STEP 8: Plot using base R map + transparent polygon overlays

```{r}
# Allow drawing in margins (optional)
par(xpd = NA)

# Plot USA map (outline and filled states)
map("usa")
map(add = TRUE, col = "grey90", fill = TRUE)

# Overlay SDM ranges
plot(st_geometry(poly_pipiens_ll), col = rgb(0, 0, 1, 0.3), border = NA, add = TRUE)       # Blue
plot(st_geometry(poly_quinque_ll), col = rgb(0, 1, 0, 0.3), border = NA, add = TRUE)        # Green
plot(st_geometry(overlap_ll), col = adjustcolor("#fc8d59", alpha.f = 1), border = NA, add = TRUE)  # Orange overlap

# Add state outlines **after everything else**
map("state", add = TRUE, col = "black", lwd = 0.5)  # State borders on top
map("usa", add = TRUE, col = "black", lwd = 1.2)  # State borders on top

# Add sampling points
counts <- read.csv("../data/ThisStudy.txt", sep = "\t")
countsSp <- SpatialPoints(coords = cbind(counts$long,counts$lat))
#points(countsSp,pch=19, cex=.1)
```

# Combine nearby patches

For mosquitoes like Culex species, where multiple generations of dispersal could easily connect nearby habitat patches within tens of kilometers.

Group overlap polygons that are within 20–50 km  (e.g., 0.3° for ~30 km) of each other into ecologically meaningful zones of secondary contact.


```{r}
# 0. Reproject to equal-area CRS
overlap_ll_proj <- st_transform(overlap_ll, aea_crs)

# 1. Make valid and extract only polygon parts
overlap_valid <- st_make_valid(overlap_ll_proj)
overlap_polygons <- st_collection_extract(overlap_valid, "POLYGON")

# 2. Now (optional) cast each multipolygon to individual polygons
overlap_parts <- st_cast(overlap_polygons, "POLYGON")

# 2. Buffer outward (~25–30 km dispersal radius ≈ 25,000 meters)
buffer_dist_meters <- 25000  # 25 km
overlap_buffered <- st_buffer(overlap_parts, dist = buffer_dist_meters)
overlap_buffered <- st_make_valid(overlap_buffered)  # ensure valid geometry before union

# 3. Union to merge touching/nearby polygons
overlap_combined <- st_union(overlap_buffered)
overlap_combined <- st_make_valid(overlap_combined)  # validate union output

# 4. Buffer inward to approximate original shape
overlap_combined <- st_buffer(overlap_combined, dist = -buffer_dist_meters)
overlap_combined <- st_make_valid(overlap_combined)  # final cleanup

# 5. Cast to multipolygon if needed
overlap_combined <- st_cast(overlap_combined, "MULTIPOLYGON")

# 6. Reproject back to WGS84 for plotting
overlap_combined <- st_transform(overlap_combined, 4326)

# ---- Plotting ----

# Allow drawing in margins (optional)
par(xpd = NA)

# Plot USA map (outline and filled states)
map("usa")
map(add = TRUE, col = "grey90", fill = TRUE)

# Overlay SDM ranges
plot(st_geometry(poly_pipiens_ll), col = rgb(0, 0, 1, 0.3), border = NA, add = TRUE)       # Blue
plot(st_geometry(poly_quinque_ll), col = rgb(0, 1, 0, 0.3), border = NA, add = TRUE)        # Green
plot(st_geometry(overlap_combined), col = adjustcolor("#fc8d59", alpha.f = 0.9), border = NA, add = TRUE)  # Orange overlap

# Add state outlines AFTER everything else
map("state", add = TRUE, col = "black", lwd = 0.5)  # State borders on top

```




# Add Kernel Density of regions of predicted overlap:

```{r}
library(MASS)
library(maps)
library(sf)

# --------------------------------------------
# Use ALL coordinates from overlap polygons
# --------------------------------------------

# Ensure valid geometry
overlap_valid <- st_make_valid(overlap_ll)

# Extract all vertex coordinates (this gives many more points than centroids)
overlap_coords <- st_coordinates(overlap_valid)

# Clean: drop any rows with NAs and duplicates
overlap_coords <- overlap_coords[complete.cases(overlap_coords), ]
overlap_coords <- unique(overlap_coords)

# Confirm number of points
cat("Number of overlap coordinates:", nrow(overlap_coords), "\n")

# --------------------------------------------
# Compute kernel density surface
# --------------------------------------------

if (nrow(overlap_coords) >= 5) {
  dens <- kde2d(overlap_coords[, 1], overlap_coords[, 2], n = 200)
}

# --------------------------------------------
# Plot map with kernel density and species ranges
# --------------------------------------------

par(xpd = NA)

# Base USA map
map("usa", fill = TRUE, col = "grey90", lwd = 0.2)

# Kernel density surface
if (exists("dens")) {
  image(dens, col = hcl.colors(20, "YlOrRd", rev = TRUE), add = TRUE)
}

# Species SDM polygons
plot(st_geometry(poly_pipiens_ll), col = rgb(0, 0, 1, 0.3), border = NA, add = TRUE)
plot(st_geometry(poly_quinque_ll), col = rgb(0, 1, 0, 0.3), border = NA, add = TRUE)
plot(st_geometry(overlap_ll), col = adjustcolor("#fc8d59", alpha.f = 0.6), border = NA, add = TRUE)

# State outlines
map("state", add = TRUE, col = "black", lwd = 0.5)
```
With reverse weighting to accentuate isolated zones of hybridization:
```{r}
library(MASS)
library(maps)
library(sf)

# --------------------------------------------
# Inverse-area weighted point sampling from overlap polygons
# --------------------------------------------

# Ensure valid geometry and cast to individual polygons
overlap_valid <- st_make_valid(overlap_ll)
overlap_parts <- st_cast(overlap_valid, "POLYGON")

# Compute area of each polygon
areas <- st_area(overlap_parts)

# Inverse weighting (normalize to sum to 1)
weights <- as.numeric(1 / areas)
weights <- weights / sum(weights)

# Choose a total number of points to sample (adjust as needed)
total_points <- 1000
n_samples <- round(weights * total_points)

# Sample points within each polygon
sampled_points <- list()
for (i in seq_along(overlap_parts)) {
  if (n_samples[i] > 0) {
    pts <- st_sample(overlap_parts[i], size = n_samples[i], type = "random")
    sampled_points[[i]] <- pts
  }
}

# Sample points within each polygon (as POINT geometries)
sampled_points <- list()
for (i in seq_along(overlap_parts)) {
  if (n_samples[i] > 0) {
    pts <- st_sample(overlap_parts[i], size = n_samples[i], type = "random")
    sampled_points[[i]] <- pts
  }
}

# Combine all sampled points into a single sfc object
all_sampled_sfc <- do.call(c, sampled_points)

# Explicitly declare this as an sfc (needed if list is empty or inconsistent)
all_sampled_sfc <- st_sfc(all_sampled_sfc, crs = st_crs(overlap_ll))

# Wrap it into an sf object
all_sampled_sf <- st_sf(geometry = all_sampled_sfc)

# Extract coordinates
sample_coords <- st_coordinates(all_sampled_sf)

# Confirm number of points
cat("Number of sampled points:", nrow(sample_coords), "\n")

# --------------------------------------------
# Compute kernel density surface from sampled points
# --------------------------------------------

if (nrow(sample_coords) >= 5) {
  dens <- kde2d(sample_coords[, 1], sample_coords[, 2], n = 200)
}

# --------------------------------------------
# Plot map with kernel density and species ranges
# --------------------------------------------

par(xpd = NA)

# Base USA map
map("usa", fill = TRUE, col = "grey90", lwd = 0.2)

# Kernel density surface
if (exists("dens")) {
  image(dens, col = hcl.colors(20, "YlOrRd", rev = TRUE), add = TRUE)
}

# Species SDM polygons
plot(st_geometry(poly_pipiens_ll), col = rgb(0, 0, 1, 0.3), border = NA, add = TRUE)
plot(st_geometry(poly_quinque_ll), col = rgb(0, 1, 0, 0.3), border = NA, add = TRUE)
plot(st_geometry(overlap_ll), col = adjustcolor("#fc8d59", alpha.f = 0.6), border = NA, add = TRUE)

# State outlines
map("state", add = TRUE, col = "black", lwd = 0.5)
```
Run kernel overlap with the merged overlap:

```{r}
# Add Kernel Density of regions of predicted overlap:

library(MASS)
library(maps)
library(sf)

# --------------------------------------------
# Use ALL coordinates from the COMBINED overlap polygons
# --------------------------------------------

# Ensure valid geometry
overlap_valid <- st_make_valid(overlap_combined)   # <-- use overlap_combined here!

# Extract all vertex coordinates (many more points than centroids)
overlap_coords <- st_coordinates(overlap_valid)

# Clean: drop any rows with NAs and duplicates
overlap_coords <- overlap_coords[complete.cases(overlap_coords), ]
overlap_coords <- unique(overlap_coords)

# Confirm number of points
cat("Number of overlap coordinates:", nrow(overlap_coords), "\n")

# --------------------------------------------
# Compute kernel density surface
# --------------------------------------------

if (nrow(overlap_coords) >= 5) {
  dens <- kde2d(overlap_coords[, 1], overlap_coords[, 2], n = 200)
}

# --------------------------------------------
# Plot map with kernel density and species ranges
# --------------------------------------------

par(xpd = NA)

# Base USA map
map("usa", fill = TRUE, col = "grey90", lwd = 0.2)

# Kernel density surface
if (exists("dens")) {
  image(dens, col = hcl.colors(20, "YlOrRd", rev = TRUE), add = TRUE)
}

# Species SDM polygons
plot(st_geometry(poly_pipiens_ll), col = rgb(0, 0, 1, 0.3), border = NA, add = TRUE)
plot(st_geometry(poly_quinque_ll), col = rgb(0, 1, 0, 0.3), border = NA, add = TRUE)

# Overlap regions
plot(st_geometry(overlap_combined), col = adjustcolor("#fc8d59", alpha.f = 0.6), border = NA, add = TRUE)

# State outlines
map("state", add = TRUE, col = "black", lwd = 0.5)
```

